Gravitational wave sensitivity curves

### 2014-06-30 Sources.tex
Cosmological sources

These divide into two classes, the unresolvable and the resolvable galactic binaries. The distinction between resolvable and unresolvable is detector specific; here we choose LISA. For the unresolvable binaries the box plotted corresponds to the estimate of the background due to \cite{Nelemans},
\begin{equation} h_{c}(f)= 5\times 10^{-21} \left(\frac{f}{10^{-3}\,\textrm{Hz}}\right)^{-2/3} \; . \end{equation}
This background is unchanging over the mission lifetime, therefore, as discussed in section \ref{sec:PTAgeneralproperties}, when plotting the amplitude on a sensitivity curve plot an observation time must be specified; here an observation time of one year is used. Estimates for the event rate of resolvable binaries centre around ${\cal{O}}(10^{3})$ events for eLISA. The box plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} has an ratio $h_{c}/h_{n}=50$ for eLISA at its peak sensitivity, the frequency range of the box is $\left(3\times10^{-4}-10^{-2}\right)\,\textrm{Hz}$, based on visual inspection of Monte Carlo population simulation results presented in \cite{Amaro-Seoane-et-al}.

\subsubsection{Extreme mass ratio inspirals}
EMRI events occur when a compact stellar mass object inspirals into a supermassive black hole. There is extreme uncertainty in the event rate for EMRIs due to the poorly constrained astrophysics in galactic centres; the best guess estimate is ${\cal{O}}(10)$ events per year with eLISA with SNR $\ge 20$. The box plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} has a characteristic strain of $h_{c}=3\times 10^{-20}$ at $10^{-2}\;\textrm{Hz}$ which corresponds to a $10\Msun$ BH inspiralling into a $10^{6}\Msun$ black hole at a luminosity distance of $1\,\textrm{Gpc}$. The frequency width of the box is somewhat unknown, EMRI events can occur into a black hole of any mass, and hence EMRIs can occur at any frequency. However there is substantial uncertainty in the black hole mass function, so the box in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} is drawn with a width comparable to that of the LISA sensitivity curve.

@@

\subsubsection{Supermassive black hole binaries}
The main target for PTAs is a stochastic background of GWs produced by a population of supermassive black hole binaries at cosmological distances. Supermassive black holes are known to lie at the centres of most galaxies and the black hole mergers are associated with the mergers of the host galaxies. The current best published limit for the amplitude of the stochastic background is $h_{c}=6\times 10^{-15}$ at a frequency of $f_{0}=1\,\textrm{yr}^{-1}$, \cite{Haasteren}. There is strong theoretical evidence that the actual background lies close to the current limit, \cite{imminentdetectionofgravitationalwaves} and \cite{Sesana-2012}.

Sources at higher frequencies are inspiralling faster and hence there are fewer of them per frequency bin. At a certain frequency, these sources will cease to be a background and become individually resolvable. The exact distinction between resolvable and unresolvable is detector specific. Plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} for the unresolvable background is a third of the current limit with a cut off frequency of $f=1\,\textrm{yr}^{-1}$ which is suggested by Monte Carlo population studies, \cite{SesanaVecchioColancino}. For the resolvable sources the amplitude of the current limit is plotted between $\left(3\times 10^{-9} - 3\times 10^{-7}\right)\,\textrm{Hz}$.

\subsection{Cosmological sources}
\cjm{Christopher, would you be happy to write a section on very high redshift sources of stochastic backgrounds. Stressing that these in principal can be at any frequency. And also saying that because of the lack of certainty here we don't include these sources in the figures. Maybe reference recent excitement caused by BICEP2 results.}

>>>>

These divide into two classes, the unresolvable and the resolvable galactic binaries. The distinction between resolvable and unresolvable is detector specific; here we choose LISA. For the unresolvable binaries the box plotted corresponds to the estimate of the background due to \cite{Nelemans},
\begin{equation} h_{c}(f)= 5\times 10^{-21} \left(\frac{f}{10^{-3}\,\textrm{Hz}}\right)^{-2/3} \; . \end{equation}
This background is unchanging over the mission lifetime, therefore, as discussed in section \ref{sec:PTAgeneralproperties}, when plotting the amplitude on a sensitivity curve plot an observation time must be specified; here an observation time of one year is used. Estimates for the event rate of resolvable binaries centre around ${\cal{O}}(10^{3})$ events for eLISA. The box plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} has an ratio $h_{c}/h_{n}=50$ for eLISA at its peak sensitivity, the frequency range of the box is $\left(3\times10^{-4}\right.$--$\left.10^{-2}\right)~\mathrm{Hz}$, based on visual inspection of Monte Carlo population simulation results presented in \cite{Amaro-Seoane-et-al}.

\subsubsection{Extreme mass ratio inspirals}
EMRI events occur when a compact stellar mass object inspirals into a supermassive black hole. There is extreme uncertainty in the event rate for EMRIs due to the poorly constrained astrophysics in galactic centres; the best guess estimate is ${\mathcal{O}}(10)$ events per year with eLISA with SNR $\ge 20$. The box plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} has a characteristic strain of $h_{c}=3\times 10^{-20}$ at $10^{-2}~\mathrm{Hz}$ which corresponds to a $10\Msun$ BH inspiralling into a $10^{6}\Msun$ black hole at a luminosity distance of $1~\mathrm{Gpc}$. The frequency width of the box is somewhat unknown, EMRI events can occur into a black hole of any mass, and hence EMRIs can occur at any frequency. However there is substantial uncertainty in the black hole mass function, so the box in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} is drawn with a width comparable to that of the LISA sensitivity curve.

@@

\subsubsection{Supermassive black hole binaries}
The main target for PTAs is a stochastic background of GWs produced by a population of supermassive black hole binaries at cosmological distances. Supermassive black holes are known to lie at the centres of most galaxies and the black hole mergers are associated with the mergers of the host galaxies. The current best published limit for the amplitude of the stochastic background is $h_{c} = 6\times 10^{-15}$ at a frequency of $f_{0}=1~\mathrm{yr}^{-1}$, \cite{Haasteren}. There is strong theoretical evidence that the actual background lies close to the current limit, \cite{imminentdetectionofgravitationalwaves} and \cite{Sesana-2012}.

Sources at higher frequencies are inspiralling faster and hence there are fewer of them per frequency bin. At a certain frequency, these sources will cease to be a background and become individually resolvable. The exact distinction between resolvable and unresolvable is detector specific. Plotted in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega} for the unresolvable background is a third of the current limit with a cut off frequency of $f=1~\mathrm{yr}^{-1}$ which is suggested by Monte Carlo population studies, \cite{SesanaVecchioColancino}. For the resolvable sources the amplitude of the current limit is plotted between $\left(3\times 10^{-9}\right.$--$\left.3\times 10^{-7}\right)~\mathrm{Hz}$.

\subsection{Cosmological sources}

In addition to the sources above, early Universe processes, such as inflation \citep{Grishchuk2005} or a first-order phase transition \citep{Binetruy2012}, could have created GWs. More speculatively, it has been hypothesised that cosmic strings could also be a potential source \citep{Damour2005,Binetruy2012}. These relic GWs allow us to explore energy scales far beyond those accessible by other means, providing insight into new and exotic physics. The excitement surrounding the tentative discovery by BICEP2 of the imprint of primordial GWs (generated during inflation) in the cosmic microwave background \citep{Ade2014}, and the subsequent flurry of activity, has shown the scientific potential of such cosmological GWs. These GW signals are so alluring because they probe unknown physics; this also makes them difficult to predict. Cosmological stochastic backgrounds have been predicted across a range of frequencies with considerable variation in amplitude. As a consequence of this uncertainty, although we could learn much from measuring these signals, we have not included them amongst the sources shown in figures \ref{fig:hc}, \ref{fig:S} and \ref{fig:omega}.
